---
title: 'Bios 6301: Assignment 2'
author: "Jonathan Lifferth"
output:
  pdf_document: default
  html_document: default
---

*Due Tuesday, 19 September, 1:00 PM*

50 points total.

Add your name as `author` to the file's metadata section.

Submit a single knitr file (named `homework2.rmd`) by email to marisa.h.blackman@vanderbilt.edu.
Place your R code in between the appropriate chunks for each question.
Check your output by using the `Knit HTML` button in RStudio.

1. **Working with data** In the `datasets` folder on the course GitHub repo, you will find a file called `cancer.csv`, which is a dataset in comma-separated values (csv) format. This is a large cancer incidence dataset that summarizes the incidence of different cancers for various subgroups. (18 points)

    1. Load the data set into R and make it a data frame called `cancer.df`. (2 points)

```{r}
cancer.df <- read.csv('~/R_projects/Bios6301/datasets/cancer.csv')
head(cancer.df)
```


    2. Determine the number of rows and columns in the data frame. (2)
    
```{r}
nrow(cancer.df)
ncol(cancer.df)
```


    3. Extract the names of the columns in `cancer.df`. (2)

```{r}
colnames(cancer.df)
```

    4. Report the value of the 3000th row in column 6. (2)

```{r}
cancer.df[3000,6]
```

    5. Report the contents of the 172nd row. (2)

```{r}
cancer.df[172,]
```

    6. Create a new column that is the incidence *rate* (per 100,000) for each row. The incidence rate is the `(number of cases)/(population at risk)`, which in this case means `(number of cases)/(population at risk) * 100,000`. (3)

```{r}
head(cancer.df)

round(.168, 2)
cancer.df['incidence rate'] <- cancer.df[,'incidence'] / cancer.df[,'population']

head(cancer.df)
```


    7. How many subgroups (rows) have a zero incidence rate? (2)

cancer.df[2,'incidence rate'] == 0

```{r}
zero_rate <- 0
zero_rate

for (i in 1:nrow(cancer.df)) {
  if (cancer.df[i,'incidence rate'] == 0) {
    zero_rate <- zero_rate + 1}
}
zero_rate
```


    8. Find the subgroup with the highest incidence rate.(3)
    
```{r}

highest_subgroup <- 0
highest_i <- 0
highest_rate <- 0

for (i in 1:nrow(cancer.df)) {
  if (cancer.df[i,'incidence rate'] > highest_rate) {
    highest_rate <- cancer.df[i,'incidence rate']
    highest_subgroup <- paste(cancer.df[i,'state'], cancer.df[i,'sex'], cancer.df[i,'race'])}
  
}

highest_rate
highest_subgroup

```


2. **Data types** (10 points)

    1. Create the following vector: `x <- c("5","12","7")`. Which of the following commands will produce an error message? For each command, Either explain why they should be errors, or explain the non-erroneous result. (4 points)

#            max(x)
#            sort(x)
#            sum(x)

```{r}
x <- c("5","12","7")
x
max(x)
?max()
# max(x) returned "7" because "Character versions are sorted lexicographically" by max()

sort(x)
# sort(x) returned "12" "5"  "7" because the first character of "12" is "1" -- the function does not treat 12 as an integer value which would place 12 at the end of the sorted vector

#sum(x)
# sum(x) returned "Error in sum(x) : invalid 'type' (character) of argument" because characters cannot be summed by this function
```


    2. For the next two commands, either explain their results, or why they should produce errors. (3 points)

#            y <- c("5",7,12)
#            y[2] + y[3]

```{r}
?c()
y <- c("5",7,12)
#y[2] + y[3]

# I was surprised to see the error "Error in y[2] + y[3] : non-numeric argument to binary operator" because I assumed that 7 and 12 could be added as integers.
# evidently, the type of these values became "character" when they were combined with "5"

typeof(y[3])

```


    3. For the next two commands, either explain their results, or why they should produce errors. (3 points)

            z <- data.frame(z1="5",z2=7,z3=12)
            z[1,2] + z[1,3]

```{r}
z <- data.frame(z1="5",z2=7,z3=12)
z
z[1,2] + z[1,3]
# this command did not produce an error because the values were arranged in a data.frame, allowing them to retain their numeric/integer values. Therefore, 7 and 12 were still eligible for addition.
```


3. **Data structures** Give R expressions that return the following matrices and vectors (*i.e.* do not construct them manually). (3 points each, 12 total)

    1. $(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)$
```{r}
c(seq(1:8), seq(from = 7, to = 1))
```
    
    2. $(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)$
```{r}
rep(1:5, times = 1:5)
```

    3. $\begin{pmatrix}
      0 & 1 & 1  \\
      1 & 0 & 1  \\
      1 & 1 & 0  \\
    \end{pmatrix}$
```{r}
x <- matrix(rep(1, 9), nrow=3)
x

x2 <- matrix(c(0,1,1,1,0,1,1,1,0), nrow=3)
x2
```

    4. $\begin{pmatrix}
      1 & 2 & 3 & 4 \\
      1 & 4 & 9 & 16 \\
      1 & 8 & 27 & 64  \\
      1 & 16 & 81 & 256 \\
      1 & 32 & 243 & 1024  \\
    \end{pmatrix}$
```{r}
x <- matrix(c((1:4), ((1:4) ^ 2), (1:4) ^ 3, (1:4) ^ 4, (1:4) ^ 5), nrow=5)
x
```



4. **Basic programming** (10 points)

    1. Let $h(x,n)=1+x+x^2+\ldots+x^n = \sum_{i=0}^n x^i$. Write an R program to calculate $h(x,n)$ using a `for` loop. As an example, use `x = 5` and `n = 2`. (5 points)

```{r}
x <- 5
n <- 2

# n starts at 0 and increases with each iteration

sum <- 0 
for (i in 0:n) {
  sum <- sum + x^i
#  print(sum)
#  sum <- sum + x^i
}

sum

```

    1. If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Write an R program to perform the following calculations. (5 points)

        1. Find the sum of all the multiples of 3 or 5 below 1,000. (3, [euler1])

```{r}

sum <- 0
for (i in 1:1000) {
  if (i%%3 == 0) {
  sum <- sum + i
} else if (i%%5 == 0) {
  sum <- sum + i
}
}

print(sum)
```

        1. Find the sum of all the multiples of 4 or 7 below 1,000,000. (2)

```{r}
sum <- 0
for (i in 1:1000000) {
  if (i%%4 == 0) {
  sum <- sum + i
} else if (i%%7 == 0) {
  sum <- sum + i
}
}

print(sum)
```

    1. Each new term in the Fibonacci sequence is generated by adding the previous two terms. By starting with 1 and 2, the first 10 terms will be $(1, 2, 3, 5, 8, 13, 21, 34, 55, 89)$. Write an R program to calculate the sum of the first 15 even-valued terms. (5 bonus points, [euler2])
    
```{r}
term_count <- 1
term_sum <- 2
a <- 1
b <- 2

while (term_count < 15) {
  num <- a + b
  if (num %% 2 == 0) {
    term_sum <- term_sum + num
    term_count <- term_count + 1
    a <- b
    b <- num
  } else {  
    a <- b
    b <- num
    }
}

term_sum
```


Some problems taken or inspired by projecteuler.

[euler1]: https://projecteuler.net/problem=1
[euler2]: https://projecteuler.net/problem=2
