---
author: 'Jonathan Lifferth'
title: 'Bios 6301: Assignment 7'
output: pdf_document
---

*Due Thursday, 02 November, 1:00 PM*

$5^{n=day}$ points taken off for each day late.

40 points total.

Submit a single knitr file (named `homework7.rmd`), along with a valid PDF output file. Inside the file, clearly indicate which parts of your responses go with which problems (you may use the original homework document as a template). Add your name as `author` to the file's metadata section. Raw R code/output or word processor files are not acceptable.

Failure to name file `homework7.rmd` or include author name may result in 5 points taken off.

### Question 1 ###

**21 points**

Use the following code to generate data for patients with repeated measures of A1C (a test for levels of blood glucose).

```{r}
genData <- function(n) {
    if(exists(".Random.seed", envir = .GlobalEnv)) {
        save.seed <- get(".Random.seed", envir= .GlobalEnv)
        on.exit(assign(".Random.seed", save.seed, envir = .GlobalEnv))
    } else {
        on.exit(rm(".Random.seed", envir = .GlobalEnv))
    }
    set.seed(n)
    subj <- ceiling(n / 10)
    id <- sample(subj, n, replace=TRUE)
    times <- as.integer(difftime(as.POSIXct("2005-01-01"), as.POSIXct("2000-01-01"), units='secs'))
    dt <- as.POSIXct(sample(times, n), origin='2000-01-01')
    mu <- runif(subj, 4, 10)
    a1c <- unsplit(mapply(rnorm, tabulate(id), mu, SIMPLIFY=FALSE), id)
    data.frame(id, dt, a1c)
}
x <- genData(500)
```

# Perform the following manipulations: (3 points each)

1. Order the data set by `id` and `dt`.
```{r}
head(x)
```
```{r}
x<-x[order(x[,'id'], x[,'dt']),]
head(x)
```


2. For each `id`, determine if there is more than a one year gap in between observations.  Add a new row at the one year mark, with the `a1c` value set to missing.  A two year gap would require two new rows, and so forth.
```{r}
library(lubridate)

x <- genData(500)
x_ordered<-x[order(x[,'id'], x[,'dt']),]
rownames(x_ordered) <- NULL

x <- x[order(x[,'id'], x[,'dt']),]
rownames(x) <- NULL
gap_years <- data.frame()

nrow(x)
for (i in 1:(nrow(x)-1)) {
  current_id <- x[i,'id']
  next_id <- x[i+1,'id']
  
    # only proceed if we're evaluating the same id
    if (next_id == current_id) {
        # what is the time gap between the current row and the next row?
        delta <- abs(difftime(x[i+1,'dt'], x[i,'dt'], units = 'days')[[1]])
        num_years <- trunc(delta / 365)
      
        # add new rows for each observation year gap
        if (num_years > 0) {
          for (j in 1:num_years) {
            new_row <- data.frame(id = current_id, dt = x[i,'dt']+years(j), a1c=NA)
    #        x <- rbind(x[1:i,], new_row, x[i+1:nrow(x),])
            gap_years <- rbind(gap_years, new_row)
        }
      }
    }

}

```
```{r}
nrow(gap_years)
```


```{r}
x2 <- rbind(x, gap_years)
x2<-x2[order(x2[,'id'], x2[,'dt']),]
rownames(x2) <- NULL

head(x2)
```

3. Create a new column `visit`.  For each `id`, add the visit number.  This should be 1 to `n` where `n` is the number of observations for an individual.  This should include the observations created with missing a1c values.
```{r}
x2$visit <- 0
count <- 1 # this is the default value for visit number
for (i in 1:nrow(x2)) {
  
  # we need to exclude the very first row, because it can't be compared to any previous row ids
  if (i > 1) {
    if (x2[i, 'id'] == x2[i-1, 'id']) {
      count <- count + 1
      x2[i, 'visit'] <- count
    }
    else { # this indicates the first row for a new id
      count <- 1
      x2[i, 'visit'] <- count
    }
      
  } else { # this condition should only apply to the very first row in the data.frame
    x2[i, 'visit'] <- 1
  }
}
```

4. For each `id`, replace missing values with the mean `a1c` value for that individual.
```{r}

for (i in 1:nrow(x2)) {
  if (is.na(x2[i,'a1c'])) {
    x2[i,'a1c'] <- mean(subset(x2,id == x2[i,'id'])$a1c, na.rm=TRUE)
  }
}

```

5. Print mean `a1c` for each `id`.
```{r}
for (i in unique(x2$id)) {
  print(paste('id: ', i, '   ' ,mean(subset(x2,id == i)$a1c, na.rm=TRUE)))
}
```

6. Print total number of visits for each `id`.
```{r}
for (i in unique(x2$id)) {
  print(paste('id: ', i, '   ' ,max(subset(x2,id == i)$visit)))
}
```

7. Print the observations for `id = 15`.
```{r}
subset(x2, id==15)
```

### Question 2 ###

**16 points**

Install the `lexicon` package. Load the `sw_fry_1000` vector, which contains 1,000 common words.

```{r}
#install.packages('lexicon')
library('lexicon')
```

```{r}
data('sw_fry_1000', package = 'lexicon')
head(sw_fry_1000)
length(sw_fry_1000)
```

1. Remove all non-alphabetical characters and make all characters lowercase. Save the result as `a`.

```{r}
sw_fry_1000[which(grepl("[^[:alpha:]]", sw_fry_1000, ignore.case = TRUE))]
a <- gsub("[^[:alpha:]]", "", sw_fry_1000)
head(a)
a <- sapply(a, tolower)
```

Use vector `a` for the following questions. (2 points each)

2. How many words contain the string "ar"?

```{r}
sum(grepl("ar", a))
```

3. Find a six-letter word that starts with "l" and ends with "r".

```{r}
# Create a regular expression to match the letter "l" at the beginning of the word
l_regex <- "^l"

# Create a regular expression to match the letter "r" at the end of the word
r_regex <- "r$"

indices <- which(grepl(l_regex, a) & grepl(r_regex, a))

print(length(a[indices[1]]))
print(a[indices[1]])

```

4. Return all words that start with "col" or end with "eck".

```{r}
a[which(grepl("^col", a) | grepl("eck$", a))]
```

5. Find the number of words that contain 4 or more adjacent consonants. Assume "y" is always a consonant.

```{r}
print(a[which(grepl('[bcdfghjklmnpqrstvwxyz]{4,}', a))])
print(length(which(grepl('[bcdfghjklmnpqrstvwxyz]{4,}', a))))
```

6. Return all words with a "q" that isn't followed by a "ui".

```{r}
grep("\\b\\w*q(?!ui)\\w*\\b", a, value = TRUE, perl = TRUE)
```

7. Find all words that contain a "k" followed by another letter. Run the `table` command on the first character following the first "k" of each word.

```{r}
k_words <- grep("\\b\\w*k\\w+\\b", a, value=TRUE)
k_words

first_chars <- character(0)

for (word in k_words) {
  match <- regexpr("k.", word)
  if (match != -1) {
    first_char <- substr(word, match + 1, match + 1)
    first_chars <- c(first_chars, first_char)
  }
}

# Use the table function to count the occurrences of each unique first character
char_count <- table(first_chars)

# Print the character count
print(char_count)

```

8. Remove all vowels. How many character strings are found exactly once?

```{r}
remove_vowels <- function(string) {
  return(gsub("[aeiouAEIOU]", "", string, ignore.case = TRUE))
}
no_vowels <- lapply(a, remove_vowels)
length(unique(no_vowels))

```



### Question 3 ###

**3 points**

The first argument to most functions that fit linear models are formulas.  The following example defines the response variable `death` and allows the model to incorporate all other variables as terms. `.` is used to mean all columns not otherwise in the formula.

```{r}
url <- "https://github.com/couthcommander/Bios6301/raw/master/datasets/haart.csv"
url <- "https://github.com/couthcommander/Bios6301/blob/main/datasets/haart.csv"

haart_df <- read.csv(url)[,c('death','weight','hemoglobin','cd4baseline')]
# the urls weren't loading so I used a copy of the dataset that I already had downloaded

haart_df <- read.csv('/Users/jonathanlifferth/R_projects/Bios6301/datasets/haart.csv')[,c('death','weight','hemoglobin','cd4baseline')]

coef(summary(glm(death ~ ., data=haart_df, family=binomial(logit))))
```

Now imagine running the above several times, but with a different response and data set each time.  Here's a function:

```{r}
myfun <- function(dat, response) {
  form <- as.formula(response ~ .)
  coef(summary(glm(form, data=dat, family=binomial(logit))))
}
```

Unfortunately, it doesn't work. `tryCatch` is "catching" the error so that this file can be knit to PDF.

```{r}
tryCatch(myfun(haart_df, death), error = function(e) e)
```

What do you think is going on?  Consider using `debug` to trace the problem.
```{r}
# debug(myfun(dat=haart_df, response=death))
```
Debugging revealed the following error: 
"Error in eval(predvars, data, env) : object 'death' not found"
Meaning that the death column in haart_df is not accessible in the scope of this function.

**5 bonus points**

Create a working function.

# eval() can't find death in the current scope/environment

```{r}

myfun2 <- function(dat, response) {
  form <- paste0(substitute(response), " ~.")
  coef(summary(glm(form, data=dat, family=binomial(logit))))
}
myfun2(dat=haart_df, response=death)
```

