---
author: 'Jonathan Lifferth'
title: 'Bios 6301: Assignment 8'
output: pdf_document
---

*Due Tuesday, 14 November, 1:00 PM*

$5^{n=day}$ points taken off for each day late.

30 points total.

Submit a single knitr file (named `homework8.rmd`), along with a valid PDF output file. Inside the file, clearly indicate which parts of your responses go with which problems (you may use the original homework document as a template). Add your name as `author` to the file's metadata section. Raw R code/output or word processor files are not acceptable.

Failure to name file `homework8.rmd` or include author name may result in 5 points taken off.

### Question 1 ###

**15 points**

Install the `readxl` package and run the following

```{r}
install.packages('readxl')
fn <- 'icd10.xlsx'
if(file.access(fn, mode = 4) == -1) {
    url <- "https://www.cdc.gov/nhsn/xls/icd10-pcs-pcm-nhsn-opc.xlsx"
    download.file(url, destfile = fn, mode = 'wb')
}
dat <- readxl::read_excel(fn, sheet = 2)
```

1. Show the class of `dat`. (1 point)

```{r}
class(dat)
```

2. Show the methods available for objects of the given class (if there are multiple classes, show methods for all classes). (3 points)
```{r}
for (class in strsplit(class(dat), " ")) {
  print(class)
  print(methods(class=class))
}
```

3. If you call `print(dat)`, what print method is being dispatched? (1 point)
```{r}
typeof(dat)
```
```{r}
?print.listof
```

```{r}
methods(print)[90]

```


4. Set the class of `dat` to be a data.frame. (1 point)
```{r}
dat <- as.data.frame(dat)
```

```{r}
class(dat)
```


5. If you call `print(dat)` again, what print method is being dispatched? (1 point)
```{r}
methods(print)[87]
```


Define a new generic function `nUnique` with the code below.

```{r}
nUnique <- function(x) {
    UseMethod('nUnique')
}
```

6. Write a default method for `nUnique` to count the number of unique values in an element. (2 points)
```{r}
nUnique.default <- function(x) {
  if(length(x) > 1) {
    return(length(unique(x)))
  } else {
    return(1)
  }
}
nUnique(c(1,2,3,2,2,2))
```


7. Check your function (2 points)

```{r, eval = FALSE}
nUnique(letters) # should return 26
nUnique(sample(10, 100, replace = TRUE)) # should return 10 (probably)
```

8. Write a data.frame method for `nUnique` to operate on data.frame objects.
This version should return counts for each column in a data.frame. (2 points)

```{r}
df <- data.frame(
    A = c(1, 2, 3),
    B = c(1, 3, 4),
    C = c(2, 4, 5)
)

names(df)[1]

df[,2]

nUnique.data.frame <- function(x) {
  for (i in 1:length(names(x))) {
  print(nUnique(x[,i]))
  }
}

nUnique(df)
```


9. Check your function (2 points)

```{r, eval = FALSE}
nUnique(dat)
```

### Question 2 ###

**15 points**

Programming with classes.  The following function will generate random patient information.

```{r}
makePatient <- function() {
  vowel <- grep("[aeiou]", letters)
  cons <- grep("[^aeiou]", letters)
  name <- paste(sample(LETTERS[cons], 1), sample(letters[vowel], 1), sample(letters[cons], 1), sep='')
  gender <- factor(sample(0:1, 1), levels=0:1, labels=c('female','male'))
  dob <- as.Date(sample(7500, 1), origin="1970-01-01")
  n <- sample(6, 1)
  doa <- as.Date(sample(1500, n), origin="2010-01-01")
  pulse <- round(rnorm(n, 80, 10))
  temp <- round(rnorm(n, 98.4, 0.3), 2)
  fluid <- round(runif(n), 2)
  list(name, gender, dob, doa, pulse, temp, fluid)
}
```
```{r}
makePatient()[1]
```


1. Create an S3 class `medicalRecord` for objects that are a list with the named elements `name`, `gender`, `date_of_birth`, `date_of_admission`, `pulse`, `temperature`, `fluid_intake`. Note that an individual patient may have multiple measurements for some measurements.  Set the RNG seed to `8` and create a medical record by taking the output of `makePatient`.  Print the medical record, and print the class of the medical record. (5 points)
```{r}
# make a new class
set.seed(8)
my_record <- list(name=NULL, gender=NULL, date_of_birth=NULL, date_of_admission=NULL, pulse=NULL,
          temperature=NULL, fluid_intake=NULL)
class(my_record) <- 'medicalRecord'
patient_data <- makePatient()

my_record$name <- patient_data[[1]]
my_record$gender <- as.character(patient_data[[2]])
my_record$date_of_birth <- patient_data[[3]]
my_record$date_of_admission <- patient_data[[4]]
my_record$pulse <- patient_data[[5]]
my_record$temperature <- patient_data[[6]]
my_record$fluid_intake <- patient_data[[7]]

my_record
```
```{r}
class(my_record)
```

2. Write a `medicalRecord` method for the generic function `mean`, which returns averages for pulse, temperature and fluids. Also write a `medicalRecord` method for `print`, which employs some nice formatting, perhaps arranging measurements by date, and `plot`, that generates a composite plot of measurements over time.  Call each function for the medical record created in part 1. (5 points)
```{r}
mean.medicalRecord <- function(x) {
  pulse <- mean(x$pulse)
  temp <- mean(x$temperature)
  fluids <- mean(x$fluid_intake)
  return(list(c("pulse: ", pulse), 
              c("temp: ", temp), 
              c("fluid intake: ", fluids)))
}

mean(my_record)
```

3. Create a further class for a cohort (group) of patients, and write methods for `mean` and `print` which, when applied to a cohort, apply mean or print to each patient contained in the cohort. Hint: think of this as a "container" for patients.  Reset the RNG seed to 8 and create a cohort of ten patients, then show the output for `mean` and `print`. (5 points)

```{r}
make_record <- function() {
  new_record <- list(name=NULL, gender=NULL, date_of_birth=NULL, date_of_admission=NULL, 
                     pulse=NULL, temperature=NULL, fluid_intake=NULL)
  class(new_record) <- 'medicalRecord'
  new_patient_data <- makePatient()
  
  new_record$name <- new_patient_data[[1]]
  new_record$gender <- as.character(new_patient_data[[2]])
  new_record$date_of_birth <- new_patient_data[[3]]
  new_record$date_of_admission <- new_patient_data[[4]]
  new_record$pulse <- new_patient_data[[5]]
  new_record$temperature <- new_patient_data[[6]]
  new_record$fluid_intake <- new_patient_data[[7]]
  return(new_record)
}
```
```{r}
class(make_record())
```

```{r}
set.seed(8)
my_cohort <- list()

for (i in 1:10) {
  new_record <- make_record() 
  my_cohort[[i]] <- new_record
}

class(my_cohort) <- 'cohort'
my_cohort
mean.cohort <- function(x) {
  for (patient in x) {
    print(patient$name)
    print(mean(patient))
  }
}

print.cohort <- function(x) {
  for (patient in x) {
    print(patient)
  }
}

```


```{r}
mean(my_cohort)
```
```{r}
print(my_cohort)
```

