---
author: 'Jonathan Lifferth'
title: 'Bios 6301: Assignment 5'
output: pdf_document
---

*Due Thursday, 12 October, 1:00 PM*

$5^{n=day}$ points taken off for each day late.

40 points total.

Submit a single knitr file (named `homework5.rmd`), along with a valid PDF output file. Inside the file, clearly indicate which parts of your responses go with which problems (you may use the original homework document as a template). Add your name as `author` to the file's metadata section. Raw R code/output or word processor files are not acceptable.

Failure to name file `homework5.rmd` or include author name may result in 5 points taken off.

### Question 1 ###

**15 points**

A problem with the Newton-Raphson algorithm is that it needs the derivative $f'$. If the derivative is hard to compute or does not exist, then we can use the *secant method*, which only requires that the function $f$ is continuous.

Like the Newton-Raphson method, the **secant method** is based on a linear approximation to the function $f$. Suppose that $f$ has a root at $a$. For this method we assume that we have *two* current guesses, $x_0$ and $x_1$, for the value of $a$. We will think of $x_0$ as an older guess and we want to replace the pair $x_0$, $x_1$ by the pair $x_1$, $x_2$, where $x_2$ is a new guess.

To find a good new guess x2 we first draw the straight line from $(x_0,f(x_0))$ to $(x_1,f(x_1))$, which is called a secant of the curve $y = f(x)$. Like the tangent, the secant is a linear approximation of the behavior of $y = f(x)$, in the region of the points $x_0$ and $x_1$. As the new guess we will use the x-coordinate $x_2$ of the point at which the secant crosses the x-axis.

The general form of the recurrence equation for the secant method is:

$$x_{i+1} = x_i - f(x_i)\frac{x_i - x_{i-1}}{f(x_i) - f(x_{i-1})}$$

Notice that we no longer need to know $f'$ but in return we have to provide *two* initial points, $x_0$ and $x_1$.

**Write a function that implements the secant algorithm.** Validate your program by finding the root of the function $f(x) = \cos(x) - x$. Compare its performance with the Newton-Raphson method -- which is faster, and by how much?  For this example $f'(x) = -\sin(x) - 1$.


```{r}
secant <- function(f, x0, x1, tol = 1e-6, max_iter = 100) {

# f: A function
# x0: initial guess
# x1: The second guess
# tol: error tolerance
# max_iter: The maximum number of iterations.

    for (i in 1:max_iter) {
    fx0 <- f(x0)
    fx1 <- f(x1)

    # Calculate next approximation for root.
    x2 <- x1 - fx1 * (x1 - x0) / (fx1 - fx0)

    # Check if the convergence criterion is met.
    if (abs(x2 - x1) < tol) {
      return(x2)
    }

    # Update guesses for root.
    x0 <- x1
    x1 <- x2
  }

  # root could not be found.
  return(NULL)
}

```


```{r}
# here I test my secant function on another simple function
f <- function(x) x^2 - 1
x0 <- 5
x1 <- 10
root <- secant(f, x0, x1)
root
```
```{r}
# here, I find the root of the function $f(x) = \cos(x) - x$
secant(function(x) cos(x) - x, x0, x1)
```
```{r}
# here's the Newton-Raphson algorithm

x<-10
while(abs(f <- x^3 + x^2 - 3*x -3) > 1e-8) {
  fp <- 3*x^2 + 2*x - 3
  x <- x - f / fp
}
x
```

```{r}
# Compare its performance with the Newton-Raphson method -- which is faster, and by how much?

x0 <- 5
x1 <- 10
# secant 
secant(function(x) cos(x) - x, x0, x1)

# Newton-Raphson
#x3 <- 10
#while(abs(f <- cos(x3) - x3) > 1e-8) {
#  fp <- sin(x3) - 1
#  x3 <- x3 - f / fp
#}
#x3

```
# I don't know why I'm getting the following error when I try to run the Newton-Raphson method :
# Line 93 "missing value where TRUE/FALSE needed"



### Question 2 ###

**20 points**

The game of craps is played as follows (this is simplified). First, you roll two six-sided dice; let x be the sum of the dice on the first roll. If x = 7 or 11 you win, otherwise you keep rolling until either you get x again, in which case you also win, or until you get a 7 or 11, in which case you lose.

Write a program to simulate a game of craps. You can use the following snippet of code to simulate the roll of two (fair) dice:

```{r}
play_craps <- function(seed, show_progress=TRUE) {
  set.seed(seed)
  game_over <- FALSE
  
  x <- sum(ceiling(6*runif(2)))
  if (show_progress) {
    print(c("x:", x))
  }
  if(x == 7 | x == 11) {
    print("You win!")
    game_over == TRUE
  } else {
    while(game_over == FALSE) {
      y <- sum(ceiling(6*runif(2)))
      if (show_progress) {
        print(c("y: ", y))
      }
      if (y == x) {
        print("You win!")
        game_over <- TRUE
      } else if (y == 7 | y == 11) {
        print("You lose :(")
        game_over <- TRUE
        }
    }
  }
  
  print("Game over ")
}


play_craps(seed=100, show_progress = FALSE)    
```

1. The instructor should be able to easily import and run your program (function), and obtain output that clearly shows how the game progressed.  Set the RNG seed with `set.seed(100)` and show the output of three games. (lucky 13 points)


```{r}
play_craps(100) 
play_craps(100) 
play_craps(100) 
```


1. Find a seed that will win ten straight games.  Consider adding an argument to your function that disables output. Show the output of the ten games. (7 points)

```{r}
for(i in 1:10){
  play_craps(1) 
}
```
# I heard after the fact that Cole wanted us to set the seed outside of the function but the instructions don't technically say that?

### Question 3 ###

**5 points**

This code makes a list of all functions in the base package:

```{r eval=FALSE}
objs <- mget(ls("package:base"), inherits = TRUE)
funs <- Filter(is.function, objs)
```

Using this list, write code to answer these questions.

1. Which function has the most arguments? (3 points)

```{r}
objs <- mget(ls("package:base"), inherits = TRUE)
funs <- Filter(is.function, objs)

top_fun <- ""
top_fun_len <- 0

for(i in 1:length(funs)) {
  fun_len <- length(formals(names(funs[i])))
  if (fun_len > top_fun_len) {
    top_fun <- names(funs[i])
    top_fun_len <- length(formals(names(funs[i])))
  }
}

print(top_fun_len)
```
```{r}
print(top_fun)
```


1. How many functions have no arguments? (2 points)

```{r}
no_args_count <- 0
for(i in 1:length(funs)) {
  fun_len <- length(formals(names(funs[i])))
  if (fun_len == 0) {
    no_args_count <- no_args_count + 1
  }
}
no_args_count
```


Hint: find a function that returns the arguments for a given function.
